---
title: "Survival analysis for readmission data"
author: "Nayef Ahmad - VCH Decision Support"
date: "March 10, 2019"
output: 
    html_document: 
        toc: yes
        toc_float: yes
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)

library(kableExtra)

# shortcuts: 
# > ctrl + alt + i ==> new code chunk
# > ctrl + shift + k ==> knit doc 
# \  ==> line break (include 2 spaces after)
```


```{r}
library(tidyverse)
library(survival)
library(ggfortify)  # one-line survival plots with ggplot2::autoplot.
library(ggpubr)

```


## Understanding readmission patterns
As the COO of a hospital, you are interested in readmission patterns, such as the proportion of patients who are readmitted to your hospital within 7 days of discharge. 

You have collected historical data on readmissions, and want to see if they can help you answer questions such as: 

* When do readmissions occur? For example, is a patient more likely to be readmitted on Day 3 after discharge, or on Day 5, or after Day 10, and so on. 
* What proportion of patients are readmitted, and what proportion are not? 


## Data
You have data on 15 patients. Seven of them were discharged, but not observed to have been readmitted yet. Thus, we don't really have the value of the variable `days before readmit` for these patients.  

```{r echo = FALSE}

df1.readmits <- 
    data.frame(stringsAsFactors=FALSE,
                patient_id = c(1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 9L, 10L, 11L, 12L,
                               13L, 14L, 15L),
       days.before.readmit = c(1L, 2L, 2L, 2L, 3L, 3L, 4L, 6L, 2L, 12L, 15L, 30L,
                               2L, 22L, 4L),
              not.censored = c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L,
                               0L, 0L),
                     notes = c(NA, NA, NA, NA, NA, NA, NA, NA,
                               "no observed readmit", "no observed readmit",
                               "no observed readmit", "no observed readmit",
                               "no observed readmit", "no observed readmit",
                               "no observed readmit")
    )



```

View the data: 
``` {r} 
df1.readmits %>% 
    kable() %>% 
    kable_styling(bootstrap_options = c("striped")) %>%
    scroll_box(width = "100%", height = "400px")

```

Now let's explore alternative ways of visualizing this data, and what conclusions we can draw from each. 


## Option 1: Density of time-to-readmission

The simplest thing to do is to just plot a histogram or density function of time-to-readmission, for the patients who have been observed as readmits. 

```{r}

p1.density <- 
    df1.readmits %>% 
    filter(not.censored == 1) %>% 
    
    ggplot(aes(x = days.before.readmit)) + 
    scale_x_continuous(expand = c(0,0)) +
    geom_density() 
    
p1.density
    
```


## Option 2: Survival curve 

Reference: https://rviews.rstudio.com/2017/09/25/survival-analysis-with-r/ 

Plot a Kaplan-Meier curve: 

```{r include = FALSE}
# Surv( ) builds the standard survival object: 
Surv(df1.readmits$days.before.readmit,
     df1.readmits$not.censored)

# Note that a “+” after the time in the print out of km indicates censoring.
```


```{r}
# Kaplan-Meier curve: 
km1.survival <- survfit(Surv(days.before.readmit, 
                             not.censored) ~ 1,
                        data = df1.readmits)

# summary(km1.survival)

```


```{r}
p2.survival <- autoplot(km1.survival,
                        conf.int = FALSE, 
                        firsty = 1) + 
    scale_y_continuous(limits = c(0,1)) + 
    scale_x_continuous(expand = c(0, 0), 
                       breaks = seq(0, 30, 2)) 

p2.survival

```




## What's the difference? 
```{r}
ggarrange(p1.density, 
          p2.survival, 
          nrow = 2)

```


There's a difference between these two: 

* Modeling time-to-readmission, **conditional on being readmitted**
* Modeling time-to-readmission, **without conditioning on readmission status** 

## Advangtages of the survival curve: 
* For the density plot, note that we throw away all data points that don't have a defined end date for `days to readmission`, because they still haven't been readmitted. A major advantage of the survival analysis is that it can use these data points to inform inferences about how long someone is likely to "survive" - i.e. remain outside of the hospital. For example, even though we don't know when Patient 10 will be readmitted (or if they ever will be), **we know that they have already "survived" past Day 1, and this is important information**. 
* From an operational perspective, at the time of discharge of a patient, **we don't know whether or not they will be readmitted**. Therefore, it may be more useful to examine readmission patterns across both populations - those who are readmitted, and those who are not. 
