---
title: "Poisson binomial simulation"
author: "Nayef"
date: "6/10/2020"
output: 
   html_document: 
     keep_md: yes
     code_folding: hide
     toc: true
     toc_float:
       collapsed: false
     toc_folding: false
     number_sections: true
---

<style type="text/css">

h1 { /* Header 1 */
  font-size: 28px;
  color: #0039a6;
}
h2 { /* Header 2 */
  font-size: 18px;
  color: #0039a6;
}
h3 { /* Header 3 */
  font-size: 14px;
  color: #0039a6;
}

</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Libraries 
```{r}
library(tidyverse)
```

# Overview
For a given WUC, there are $I$ tails. 

Given: prob of failure of tail $i$ in a given window, denoted $p_i$. 

We will treat each tail's status over the window as a single Bernoulli trial. The total number of failures, $T$ is a sum of Bernoulli trials, which has a [Poisson binomial distribution](https://en.wikipedia.org/wiki/Poisson_binomial_distribution)


# Data 
Let's draw the $p_i$ values from $U(0,1)$. 

## Functions

```{r}
gen_params <- function(num_tails = 15, 
                            max_fail_prob = 1.0){
  if (max_fail_prob > 1){
    stop("Max fail prob must be less than or equal to 1.0")
  }
  
  tail_id <- paste0("sn_", 1:num_tails)
  fail_prob <- runif(num_tails, max = max_fail_prob)
  
  df <- data.frame(tail_id = tail_id, 
                   fail_prob = fail_prob)
  
  return(df)
} 


gen_fails <- function(df){
  df_with_fail <- 
    df %>% 
    mutate(is_fail = map_dbl(fail_prob, 
                             function(x){
                               sample(c(1,0), 
                                      size = 1, 
                                      prob = c(x, 1-x))
                               }
                             ))
  return(df_with_fail)
  
}


total_fails <- function(df){
  sum(df$is_fail)
}


gen_totals <- function(df_with_params, iter = 1000){
  totals <- replicate(iter, 
                      df_with_params %>% gen_fails() %>% total_fails())
  
  return(totals)
  
}


calc_mean_sd <- function(df_with_params){
  df <- 
    df_with_params %>% 
    mutate(not_fail_prob = 1-fail_prob, 
           pq_product = fail_prob * not_fail_prob)
  
  mean_sd_list <- list(mean = sum(df$fail_prob),
                   sd = sqrt(sum(df$pq_product)))
  return(mean_sd_list)
}


# fn to plot with overlaid normal
plot_pois_binom <- function(num_tails = 15, 
                            max_fail_prob = 1.0, 
                            binwidth = 5, 
                            iter = 1000){
  # generate params:  
  df_with_params <- 
    gen_params(num_tails = num_tails, 
               max_fail_prob = max_fail_prob) 
  
  # generate distribution of pois binom: 
  totals_vector <- 
      gen_totals(df_with_params)
  
  # get mean, sd for normal overlay on plot: 
  pois_binom_mean <- calc_mean_sd(df_with_params)[[1]] %>% round(1)
  pois_binom_sd   <- calc_mean_sd(df_with_params)[[2]] %>% round(1)
  
  # plot
  data.frame(total_fails = totals_vector) %>% 
    ggplot(aes(x = total_fails)) + 
    geom_histogram(aes(y=..density..), 
                   binwidth = binwidth) + 
    # stat_density(geom="line", 
    #              col = "black") +
    
    labs(title = str_glue("Distribution of the sum of {num_tails} Bernoulli trials (aka Poisson binomial distribution)"), 
         subtitle = str_glue("Num tails: {num_tails}  \nFail probabilities: U(0, {max_fail_prob})  \n\nEstimated mean: {pois_binom_mean}  \nEstimated SD: {pois_binom_sd}  \n\nRed line is Normal dist with matching mean, sd"), 
         caption = str_glue("Histogram binwidth: {binwidth}\nIterations: {iter}")) + 
    
    stat_function(fun = dnorm, 
                  color = "red",
                  size = 1, 
                  args = list(mean = pois_binom_mean, 
                              sd = pois_binom_sd))
}



```


# Case 1: Small sample 
```{r}
plot_pois_binom(num_tails = 15, max_fail_prob = 1, binwidth = 1)
```


# Case 2: Large sample
```{r}
plot_pois_binom(num_tails = 300, max_fail_prob = 1, binwidth = 5)
```

